---
title: "Statistical and Math Review"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
runtime: shiny    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(PSUEC469) # Other libraries may be required.


set.seed(458) # make random results reproducible

```

# Summation Operator

Property 1

$$\sum_{i=1}^{n}c = nc$$
Property 2
$$\sum_{i=1}^{n}cx_i = c\sum_{i=1}^{n}x_i$$
Property 3
$$\sum_{i=1}^{n} (ax_i + b y_i) = a\sum_{i=1}^{n}x_i+ b\sum_{i=1}^{n}y_i$$

## Questions

- What is $\sum_{m=1}^{4} 8k -6m$?

$$\sum_{m=1}^{4} 8k -6m = \\$$

- Summation notation for $0 + .25 + .5 + .75$



- If the mean is $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i$ what is $\frac{1}{n} \sum_{i=1}^n (2x_i - \bar{x})$?

$$\frac{1}{n} \sum_{i=1}^n (2x_i - \bar{x}) = \\$$



- If the variance is defined as $\frac{1}{n} \sum_{i=1}^n (x_i - \bar{x})^2$ what is the variance of 2x?

#  Logs

We will spend a lot of time in this class with the equations that are linear in the parameters but that can include polynomials, logs and other transformations like that.

A function that is linear in the parameters are additively separable where the parameters are just multiplied by some function of other variables but no other parameters.

These are all linear in the parameters:

+ $y = \beta_1 + \beta_2 x_1 + \beta_3 x_2$
+ $y = \beta_1 + \beta_2 x^2 + \beta_3 x^3$
+ $y = \beta_1 + \beta_2 \ln(x_1) + \beta_3 x_2$
+ $y = \beta_1 + \beta_2 x_1 + \beta_3 x_2x_1$

These are not:

+ $y = \frac{\beta_1}{\beta_2} +  \beta_1 x_1 + \beta_2 x_2$
+ $y = \beta_1 \beta_2 x_1 + \beta_1 x_2 + \beta_2 x_1$
+ $y = \beta_1 + x_1^{\beta_2}  + \beta_3 x_2$

We will often transform equations, by taking the natural log, which we just call the log, or square root to turn them into linear equations.


Logarithm rules:

+ $\ln{x y} = \ln{x} + \ln{y}$
+ $\ln{\frac{x}{y}} = \ln{x} - \ln{y}$
+ $\ln{x^y} = y\ln{x}$
+ $f(x) = \ln{x} \Rightarrow f'(x) = \frac{1}{x}$
+ $\lim_{x \rightarrow 0^+} = -\infty $
+ $\ln(1)=0$
+ Undefined at zero and complex valued when less than zero, $\ln(-1) = i\pi$.  We don't use complex values in this class.

Our equations will generally be polynomials and we will often be interested in the _marginal effects_ of a change in some variable.  This generally means taking the derivative with respect to that variable. I will keep things down to some simple cases but you should review these basic derivatives and the chain rule. I will use both the Lagrange, $f'$, and Leibniz, $\frac{d}{dx}$, notation interchangibly.

+ $\frac{d}{dx} C = 0$
+ $\frac{d}{dx} x^2 = 2 x$
+ $\frac{d}{dx} x^n = n x^{n-1}$
+ $\frac{d}{dx} e^x = e^x$
+ $\frac{d}{dx} e^{kx} = ke^{kx}$
+ $\frac{d}{dx} \ln{x} = \frac{1}{x}$
+ Chain rule $\frac{dz}{dx} = \frac{dz}{dy} \frac{dy}{dx}$


## Questions

- $\ln(y) = \beta_1 z_1 + \beta z_2^2$ Is this linear in the parameters? Can it be linearized with a ln transform? If so, show.

- $y = \beta_1\beta_2x^{\beta_3}$ Is this linear in the parameters? Can it be linearized with a ln transform?  If so, show.


- Given $Income = 10 + .5 ~experience - .1 ~experince^2$, which explains income with years of experience, what is the marginal effect of a year of experience? 

- Given $Income = 10 + .5 ~experience - .1 ~experince^2$, which explains income with years of experience, when will income be maximized? 

- Given a demand function, $Q = e^{-5p} + 4$ what is the marginal effect of price changes?

- Given a demand function, $Q = e^{-5p} + 4$ at what price will people stop buying the good?

- Given a demand function, $Q = e^{p^2-5p} + 4$ what is the marginal effect of price changes?

# Probability Discrete and Continuous


```{r}

library(MASS)

Sigma <- matrix(c(10,3,3,2), 2,2)
mvrnorm(n = 20 , mu = c(.5, .5), Sigma )

```

